 Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings

### 2020-07-17 paper.Rmd
update examples

The next section steps through three use cases of our approach using simulated
and real data. In the first case study, we step through three simulated datasets
to show the `liminal` interface. The first simulation contains 3 clusters
sampled from a 5-d gaussian distribution that are well embedded by t-SNE.
The second is another example of spherical clusters but this time, they are
tightly clustered and are distorted by t-SNE, which is immediately apparent in 
the tour view. In the final simulation we
look at a tree structure (similar to trajectories in single cell 
transcriptomics) that
highlights an iterative approach to modifying t-SNE views based on what we
learn in the high-dimensional space. We show how through simple linked
brushing, we can see when the assumptions of continuity of the embedding map
are broken.

@@

### Case Study 1: Exploring tree structured data with tours and t-SNE

In this case study we explore some simulated noisy tree structured data (figure
\@ref(fig:fake-trees)), our interest here is how t-SNE visualisations break
topology of the data, and then seeing if we can resolve this by tweaking the
default parameters with reference to the global view of the data set.

@@

<!-- can we show R code? -->

@@

First, we downloaded the raw mouse retinal single cell RNA-seq data from
@Macosko2015-ot using the `scRNAseq` Bioconductor package [@scRNAseq-d]. We
have followed a standard workflow for pre-processing and normalizing this data 
(described by @Amezquita2020-at): we performed QC using the the `scater`
by removing cells with high proportion of mitochondrial gene expression and low
numbers of genes detected, we log-transformed and normalised the expression 
values and  finally selected highly variable genes (HVGs) using `scran` 
[@McCarthy2017; @Lun2016]. The top ten percent of HVGs were used to subset
the normalised expression matrix and compute PCA using the first 25 components.
Using the PCs we built a shared nearest neighbours graph (with $k = 10$) and 
used Louvain clustering to generate clusters. To check and verify the 
clustering we construct a liminal view. Because most off the variability in
expression is explained by the first 5 PCs, we tour those. The t-SNE view
is computed from all PCs

>>>>

The next section steps through three use cases of our approach using simulated
and real data.  The videos presented show the visual 
appearance of the `liminal` interface, and how we can interact with the tour
via the controls previously described.
In the first case study, we look at three different examples
where t-SNE is applied to three simulations where cluster structure and geometry
is known with varying success.

@@

### Case Study 1: Exploring simulation data with tours and t-SNE

To begin we look at simulated datasets that reproduce known facts
aboutthe  t-SNE algorithm. Our first data set consists of five spherical 5-$d$
Gaussian clusters embedded in 10-$d$ space, each cluster has the same 
covariance matrix. We then computed a t-SNE layout using default settings,
and set up the `liminal` linked interface with grand tour on the 10-$d$
observations (video - ).

<embedded video liminal gaussian>

We see in the video that t-SNE has correctly split out each cluster and
laid them out in a star like formation. This is agrees with the tour view,
where once we start the animation, the five clusters begin to appear but
generally points are more concentrated in the projection view compared to 
the tSNE layout. This can be seen from our brushing action on the tSNE side.

<embedded video liminal hierachcial >

We repeat the process of simulating Gaussian clusters in video <>, but this time
we have a cluster that has more granular structure. This time we have two
Gaussian clusters with equal covariance embedded in $10-d$, and a third cluster
with hierachcial structure. This cluster has two 3-$d$ clusters embedded in 
10-$d$, where the second cluster is subdivided into three smaller clusters, that are each equidistant from each other and have the same covariance structure.
Again tSNE has correctly identified the sub-clusters but their relative 
locations to each other is distorted, with the orange and blue groups being
far from each other in the tour view. We see in this case that is difficult
to see the sub-clusters in the tour view, however, once we zoom and highlight
they become more apparent. When we brush the sub-clusters in the tSNE, their
relative placement is again exaggerated, with the tour showing that they
are indeed much closer than the impression the tSNE view gives.

Next explore some simulated noisy tree structured data (figure
\@ref(fig:fake-trees)), our interest here is how t-SNE visualisations break
topology of the data, and then seeing if we can resolve this by tweaking the
default parameters with reference to the global view of the data set.

@@

@@

First, we downloaded the raw mouse retinal single cell RNA-seq data from
@Macosko2015-ot using the `scRNAseq` Bioconductor package [@scRNAseq-d]. We
have followed a standard workflow for pre-processing and normalizing this data 
(described by @Amezquita2020-at): we performed QC using the the `scater`
by removing cells with high proportion of mitochondrial gene expression and low
numbers of genes detected, we log-transformed and normalised the expression 
values and  finally selected highly variable genes (HVGs) using `scran` 
[@McCarthy2017; @Lun2016]. The top ten percent of HVGs were used to subset
the normalised expression matrix and compute PCA using the first 25 components.
Using the PCs we built a shared nearest neighbours graph (with $k = 10$) and 
used Louvain clustering to generate clusters. To check and verify the 
clustering we construct a liminal view. Because most off the variability in
expression is explained by the first 5 PCs, we tour those. The t-SNE view
is computed from all PCs.
